The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-22 paper.qmd
finished the Visualising uncertainty that is not uncertainty section

# Visualising Uncertainty that is not uncertainty
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations. Some authors do not describe uncertainty as a means of transparency, but rather as a variable of importance in of itself [@Blenkinsop2000]. Other authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. While uncertainty can provide useful as its own variable of interest, or in decision making, it is important to recognize the "uncertainty" in these cases are not acting as "uncertainty" at all. The "uncertainty" in these cases is acting as signal. This is easiest to understand with an example.

@@

If this stipulation is true, then we would expect the visualisation methods and design principles of "normal" information visualisation to apply to these "uncertainty" visualisations. This is exactly what the literature shows.

Experiments that evaluate graphics on how useful they are for value extraction and decision making rarely consider how difficult this "true" statistic is to extract from the chart. Graphics are compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020]  even though the bounds of that class is rife with disagreement. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty even though histograms are frequently discussed as an uncertainty visualisation.

When we view these evaluation studies through the lens of "How do you *actually* extract the *statistic required to answer the question*?" the results often become obvious, or easily explained using existing work in information visualisation and statistics. @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function). @Hullman2015 found that a visualisation that allows viewers to estimate a probability through counting outperformed one that required a complicated area calculation.@Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. The entire uncertainty visualisation literature is awash with examples like this.
*****

@fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty. A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map. While the input into the graphic has changed, graphics, just like statistics, are not defined by their input data.

@@

There seems to be a divide between those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate. The former view uncertainty as signal (i.e. as a variable of interest on its own, or as required information for decision making) and the later view uncertainty as noise (i.e. as a variable needed to suppress the signal in the plot).

Of course, there is nothing wrong with visualizing variance, error, bias, and any other statistic used to depict uncertainty as a signal. Just like any other statistic, these metrics provide important and useful information for analysis and decisions. However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation, they are acting an estimate and they should be treated as such. For these reasons decision making and visualisations that explicitly depict uncertainty measures will not be the focus of this review. The following sections will investigate several approaches to uncertainty visualisation and discuss how successful each approach is in fulfilling the signal-suppression criteria.

>>>>

## Visualising uncertainty that is not uncertainty
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations. Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all. It is acting as signal. This is easiest to understand with an example.

@@

Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000]. This group are *also* viewing uncertainty as a signal, albeit more explicitly than the decision making group. If the purpose of uncertainty in these cases is to convey information rather than obfuscate it, then there is no reason to assume these "uncertainty visualisations" would behave any differently to normal information visualisations. This is exactly what the literature shows.

Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat. Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease. This is the logic that results in predictable results. @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function). @Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. @Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. These experiments all have the same fundamental finding, if we create a problem by obfuscating information, then the representation that makes the obfuscated information easier to extract outperforms those that don't. This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks. We have known perceptual tasks have more noise than text representations for a long time [@Cleveland1984] and it has little to do with uncertainty.

Additionally, if uncertainty visualisation for decision making is a "wrong information" problem, then we would expect uncertainty visualisations made for this purpose to simply adjust the statistic they display of rather than a change the visual elements of the graphic. This is precicely what we observe. @fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map. While the input into the graphic has changed, graphics, just like statistics, are not defined by their input data.

This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. This disagreement comes from two different conceptualisations of what an uncertainty visualisation *is*. There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate. The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.

@@

The easily explained results in evaluation experiments, and the exceedance probabiltiy map show the problem of viewing uncertainty as a signal instead of noise. If we view the purpose of viewing uncertainty visualisation to simply extract a probability or variance (reguardless of whether or not it is for value extraction or decision making) then the best performing visualsiation will be one that makes that statistic easy to extract. Even better is to avoid displaying a visualisation all together and simply provide the probability or variance with text.

Of course, we do not beleive there is anything wrong with explicitly visualizing variance, error, bias, or any other statistic used to depict uncertainty as a signal. Just like any other statistic, these metrics provide important and useful information for analysis and decisions. However, there is no interesting visualisation challenge associated with these graphics, and they do not require any special visualisation, they are acting an estimate and they should be treated as such.

For these reasons decision making and visualisations that explicitly depict uncertainty measures will not be the focus of this review. We will instead focus on the perspective that uncertainty visualisation serves to obfuscate signal, and should be considered a function that can be *applied* to any graphic to suppress false signals. The following sections will investigate several approaches to uncertainty visualisation and discuss how successful each approach is in fulfilling the signal-suppression criteria.
